Scaling through abstractions – high-performance vectorial wave simulations for seismic inversion with Devito

### 2020-06-02 response.md

> The computational kernels have really high arithmetic intensity that is great for the domain, but not as challenging for optimization

High arithmetic intensity does not imply "non challenging optimization". In fact, it's the exact opposite, and there are several examples in computational science that we could use as evidence, even outside of seismic. Take high-order finite (spectral) elements for instance -- you may get an extremely good arithmetic intensity, but without tensor-product basis functions and suitable optimizations such as sum-factorization you'll likely end up with a notable, yet totally misleasing, GFlops/s performance, as a huge fraction of flops will stem from redundant computation (e.g., common sub-expressions, lack of code hoisting, distributivity). The same story happens with our TTI. As we explain in the paper [section ...], without capturing the cross-iteration common sub-expressions (due to high order derivatives), we would have obtained remarkable GFlops/s performance, but terrible GPoints/s. And capturing such redundancies is indeed quite challenging. In fact, this is one of the reasons several companies are today so much interested in Devito: machines (compilers) can be better than humans at detecting redundancies, and definitely way faster than them.

> The parallelism applied so far is straight up MPI which is a bit behind times these days.

In 2020 MPI is still de facto a widely adopted approach for domain decomposition.

## R3

> It was a little confusing but the MPI parallelism appears to essentially be a domain decomposition style parallelism.

As we elaborate in R2 (with references to the paper), it's not just about decomposing a grid. Devito deals with distribution of sparse functions, python-level distributed arrays, decomposition over sub-domains and much more.

> There was some mention of threads, in particular there seems to be several compilation passes that abstracts out the parallelism which can include SIMD, shared-memory and MPI. Only the MPI is detailed.

Correct. We refer to other papers for all these other aspects.

@@

See reply to R1 -- basically, (i) lack of resources and (ii) Devito not ready yet for extreme scale at the time we ran the experiments.

> It appears much of the performance result are provided in the references so it is unclear as to what the purpose of this paper is for an SC audience.

We are showing performance results of three different codes -- TTI, elastic, and isotropic acoustic -- thus exploring different physics and discretizations. These results are, respectively, in sections [...]. As we explain in our reply to R2, ours is an application-oriented paper built off high-level abstractions running on a cloud-based system. To achieve this, we're using several bleeding edge technologies -- an established DSL-based system such as Devito (used in companies and academia), the Azure cloud, Docker (for ease of installation, use and reproducibility), the rapidly-spreading GitHub Actions for continuos integration, and much more. The use and orchestration of these technologies is thouroughly described in the paper, and this should be, in our opinion, of strong interest for the SC audience.

> For actual seismic surveys, the imaging (interpretation) part is only 10% of the survey. [...] while important doesn't address the real "industrial" problems, it helps a little.

@@

The industrial problem does indeed involve muliple parts, such as data acquisition, processing, and, most importantly, velocity model building. We do not claim here to solve all these problems; however, we demonstrate that the combination of advanced technolgies such as high-level DSLs and compilers is key to highly efficient sesimic modeling and imaging at industrial scale. The complexity of model building requires advanced optimization methods. The straightforward access to computationally and easy-to-develop wave-equation solvers enable research and development for seismic inversion. Devito is therfore used for the development of inital background model building [@mojica2019tab] or cycle skipping mitigating methods [@rizzuti2020EAGEtwri; @louboutin2020SEGtwri] and compressive least square migration [@witte2018cls] based on a linear algebra framework for seismic inversion built on top of devito [@witte2018alf]. We have added a clarification in the introduction.

> Dropping the discussion on the 3D elastic wave and [...] might demonstrate the value of Devito better.

We believe that showing the new capabilities of Devito -- the tensor language -- for a different physics and discretization is as important to the SC audience as a strong scaling experiment. In fact, strong scaling may, to some extent, be regarded as more of an academic exercise in seismic imaging:

* even with high-frequency FWI, a typical shot-level computation requires no more than a bunch of nodes (often order of units);
* there is an outer-level of parallelism for inversion, essentially a farm (master-slave) where the individual workers and internally MPI-parallel.

## R4 
Detailed Comments for Authors
    This manuscript describes optimizations for Devito, a symbolic DSL designed for geophysics, to extend its use for industrial scale modeling of seismic inversion problems.
    Strengths:
        added support for vector and tensor objects for staggered-grid finite-differences
        demonstrated solving large-scale inverse problems (requiring large memory and computation)
        demonstrated its use on Cloud-based HPC
        paper is well-written
    Weaknesses:
        One issue is that I cannot tell how much of this work is new
        contributions. The paper reads a lot like an overview of previous work in many parts. The parts that I am sure are new work (e.g., adding support for vector and tensors), I am unclear of the amount of effort or innovation involved.
        I like this paper and the work (the ease-of-use of Devito is impressive), but I think for SC we would need more information on its scalability and performance.
Comments for Revision
        Please address the weaknesses above
        Is section III describing new work? It is unclear to me if this is an overview or new work for this paper…
        Is the parallelism just via MPI? What are the possibilities for this code in terms of heterogeneous clusters? Will it be able to take advantage of GPUs?
    Minor issues:
    (1) section 3, 1st paragraph: "must to be" => "must be"
    (2) section 1, 3rd paragraph: make an itemized list?
Scored Review Questions WEAK REJECT (2)

### 2020-06-02 response.md

* even with high-frequency FWI, a typical shot-level computation requires no more than a bunch of nodes (often order of units);
* there is an outer-level of parallelism for inversion, essentially a farm (master-slave) where the individual workers and internally MPI-parallel.

## R4

* One issue is that I cannot tell how much of this work is new

We have improved the manuscript to clarify what the contributions are. These are also now reiterated throughout the whole paper. TODO -- need to add pointers to what we added?

* [...] vector and tensors [...] I am unclear of the amount of effort or innovation involved.

The number of additions to the codebase is notable: https://github.com/devitocodes/devito/pull/873
And for Devito and his users, this was an invaluable addition.

@@

* but I think for SC we would need more information on its scalability and performance.

The single-node performance is discussed in other (cited) articles. In the previous responses (R1, R2, R3), we have elaborated why the lack of scalability experiments.

@@

* Is section III describing new work? It is unclear to me if this is an overview or new work for this paper…

Yes. We have improved the text to make it more explicit and detailed.

@@

* Is the parallelism just via MPI?

The inversion (outer-level of parallelism) requires what is essentially task parallelism; each task is internally MPI-parallel (domain decomposition). In this context task-parallelism means paralleization over sources, which we have accomplished using `batch-shipyard` in a serverless settting (see Section TODO). We remark that this allows optimal usage of the typical cloud infrastructure, where one pays based on actual usage of computational resources.

@@

* What are the possibilities for this code in terms of heterogeneous clusters? Will it be able to take advantage of GPUs?

GPU support is in development and currently (Devito v4.2) supported for some applications. Its capabilities and related applications will be fully covered in a future dedicated paper.

* Minor issues [...]

Fixed, thanks.

### 2020-06-02 response.md

> There was some mention of threads, in particular there seems to be several compilation passes that abstracts out the parallelism which can include SIMD, shared-memory and MPI. Only the MPI is detailed.

Correct. We refer to other papers for all these other aspects.

> One question that arises, in the performance is how does Devito account for FMA (Fused Multiple Adds)?

The operations determining the performance of a code are the source level ones, while FMA is a lower level concept (ie, a scalar FMA would account for two operations). This is not Devito, but rather the model adopted everywhere, otherwise two performance values from two different architectures, one with FMAs the other without, would not even be comparable.

@@

The industrial problem does indeed involve muliple parts, such as data acquisition, processing, and, most importantly, velocity model building. We do not claim here to solve all these problems; however, we demonstrate that the combination of advanced technolgies such as high-level DSLs and compilers is key to highly efficient sesimic modeling and imaging at industrial scale. The complexity of model building requires advanced optimization methods. The straightforward access to computationally and easy-to-develop wave-equation solvers enable research and development for seismic inversion. Devito is therfore used for the development of inital background model building [@mojica2019tab] or cycle skipping mitigating methods [@rizzuti2020EAGEtwri; @louboutin2020SEGtwri] and compressive least square migration [@witte2018cls] based on a linear algebra framework for seismic inversion built on top of devito [@witte2018alf]. We have added a clarification in the introduction.

> Dropping the discussion on the 3D elastic wave and [...] might demonstrate the value of Devito better.

We believe that showing the new capabilities of Devito -- the tensor language -- for a different physics and discretization is as important to the SC audience as a strong scaling experiment. In fact, strong scaling may, to some extent, be regarded as more of an academic exercise in seismic imaging:

* even with high-frequency FWI, a typical shot-level computation requires no more than a bunch of nodes (often order of units);
* there is an outer-level of parallelism for inversion, essentially a task farm where the individual workers is internally MPI-parallel.

@@

## R4

> One issue is that I cannot tell how much of this work is new

@@

We have improved the manuscript to clarify what the contributions are. These are also now reiterated throughout the whole paper. TODO -- need to add pointers to what we added?

> [...] vector and tensors [...] I am unclear of the amount of effort or innovation involved.

@@

The number of additions to the codebase is notable: https://github.com/devitocodes/devito/pull/873
And for Devito and his users, this was an invaluable addition.

> but I think for SC we would need more information on its scalability and performance.

@@

The single-node performance is discussed in other (cited) articles. In the previous responses (R1, R2, R3), we have elaborated why the lack of scalability experiments.

> Is section III describing new work? It is unclear to me if this is an overview or new work for this paper…

Yes. We have improved the text to make it more explicit and detailed.

> Is the parallelism just via MPI?

The inversion (outer-level of parallelism) requires what is essentially task parallelism; each task is internally MPI-parallel (domain decomposition). In this context task-parallelism means paralleization over sources, which we have accomplished using `batch-shipyard` in a serverless setting (see Section II end). We remark that this allows optimal usage of the typical cloud infrastructure, where one pays based on actual usage of computational resources.

> What are the possibilities for this code in terms of heterogeneous clusters? Will it be able to take advantage of GPUs?

### 2020-06-02 paper.tex
more rephrasing

\usepackage{color}
\usepackage{fancyvrb}
\usepackage{fvextra}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}, breaklines=true}
% Add ',fontsize=\small' for more characters per line
\newenvironment{Shaded}{}{}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.00,0.44,0.13}{\textbf{{#1}}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.56,0.13,0.00}{{#1}}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.25,0.63,0.44}{{#1}}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.25,0.63,0.44}{{#1}}}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.25,0.63,0.44}{{#1}}}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.25,0.44,0.63}{{#1}}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.25,0.44,0.63}{{#1}}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textit{{#1}}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.00,0.44,0.13}{{#1}}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{1.00,0.00,0.00}{\textbf{{#1}}}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.02,0.16,0.49}{{#1}}}
\newcommand{\RegionMarkerTok}[1]{{#1}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{1.00,0.00,0.00}{\textbf{{#1}}}}
\newcommand{\NormalTok}[1]{{#1}}
\newcommand{\devito}{\href{https://github.com/devitocodes/devito}{Devito} }

@@

This paper is organized as follows: first, we provide a brief overview of
Devito and its symbolic API and present the distributed memory
implementation that allows large-scale modelling and inversion by means of
domain decomposition. We then provide a brief comparison with a state of
the art hand-coded wave propagator to validate the performance previously
benchmarked with the roofline model (\cite{patterson, devito-compiler,
devito-api, louboutin2016ppf}). \textcolor{red}{Before concluding, results from the
Cloud-based experiments introduced above are presented}.

@@

We here provide a succinct description of distributed-memory parallelism
in Devito; the interested reader should refer to the MPI tutorial at
\href{https://github.com/devitocodes/devito/blob/v4.2/examples/mpi/overview.ipynb}{mpi-notebook}
for thorough explanations and practical examples.
\textcolor{red}{We highlugh the main feature of the distributed framework tha enable the industry scale
applications we describe in this paper. A more thorough study of its implementation, including `mpi4numpy' and
a complete scaling analysis will be presented in a later paper. Initial results assciated with early version of \devito's
distributed computing capabilities can be found in \cite{luporini2019adp} on a conventional cluster and 
in \cite{witte2019ecl} that contains results on AWS.}

@@

While the subsurface image obtained in section \ref{d-imaging-example-on-azure}
utilized anisotropic propagators capable of mimicking intricate physics, in order
to model both the wave kinematics and amplitudes correctly, elastic
propagators are required. These propagators are, for example, extremely
important in global seismology since shear surface waves (which are
ignored in acoustic models) are the most hazardous. In this section, we
exploit the tensor algebra language introduced in \devito v4.0 to express
an elastic model with compact and elegant notation.

>>>>

\usepackage[normalem]{ulem}
\usepackage{color}
\usepackage{fancyvrb}
\usepackage{fvextra}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}, breaklines=true}
% Add ',fontsize=\small' for more characters per line
\newenvironment{Shaded}{}{}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.00,0.44,0.13}{\textbf{{#1}}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.56,0.13,0.00}{{#1}}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.25,0.63,0.44}{{#1}}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.25,0.63,0.44}{{#1}}}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.25,0.63,0.44}{{#1}}}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.25,0.44,0.63}{{#1}}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.25,0.44,0.63}{{#1}}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textit{{#1}}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.00,0.44,0.13}{{#1}}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{1.00,0.00,0.00}{\textbf{{#1}}}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.02,0.16,0.49}{{#1}}}
\newcommand{\RegionMarkerTok}[1]{{#1}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{1.00,0.00,0.00}{\textbf{{#1}}}}
\newcommand{\NormalTok}[1]{{#1}}
\newcommand{\devito}{\href{https://github.com/devitocodes/devito}{Devito} }

@@

This paper is organized as follows: first, we provide a brief overview of
Devito and its symbolic API and present the \textcolor{red}{newly developped} distributed memory
implementation that allows large-scale modelling and inversion by means of
domain decomposition. We then provide a brief comparison with a state of
the art hand-coded wave propagator to validate the performance previously
benchmarked with the roofline model (\cite{patterson, devito-compiler,
devito-api, louboutin2016ppf}). \textcolor{red}{Before concluding, results from the
Cloud-based experiments introduced above are presented}.

@@

We here provide a succinct description of distributed-memory parallelism
in Devito; the interested reader should refer to the MPI tutorial at
\href{https://github.com/devitocodes/devito/blob/v4.2/examples/mpi/overview.ipynb}{mpi-notebook}
for thorough explanations and practical examples.
\textcolor{red}{We highlight the main features of the distributed framework tha enable the industry scale
applications we describe in this paper. A more thorough study of its implementation, including `mpi4numpy' and
a complete scaling analysis will be presented in a later paper. Initial results assciated with early version of \devito's
distributed computing capabilities can be found in \cite{luporini2019adp} on a conventional cluster and 
in \cite{witte2019ecl} that contains results on AWS.}

@@

While the subsurface image obtained in section \ref{d-imaging-example-on-azure}
utilized anisotropic propagators capable of mimicking intricate physics, in order
to model both the wave kinematics and amplitudes correctly, elastic
propagators are required. These propagators are, for example, extremely
important in global seismology since shear surface waves (which are
ignored in acoustic models) are the most hazardous. \textcolor{red}{In this section, we 
introduce the new tensor algebra language implemented in \devito that enables the simple and 
high-level definition of vectorial and tensorial physics. This API extension also implements 
fully automatized staggered grid and staggered ginite-difference generally required for a stable discretization
of vectorial PDEs. This extension allows us to express an elastic model with compact and elegant notation 
and model its solution computationally efficiently with the \devito compiler}

\sout{\textcolor{green}{In this section, we
exploit the tensor algebra language introduced in \devito v4.0 to express
an elastic model with compact and elegant notation.}}
